\specialhead{ACKNOWLEDGMENT}

First and foremost, I would like to thank my advisor,
Prof. Boleslaw K. Szymanski, for his guidance throughout my tenure
as a graduate student. It was a great fortune to work with him during my doctoral study. This work would not have been possible without his guidance and his commitment to excellence in research. I learned from him not only the fundamentals of conducting scientific research but also a hardworking and passionate attitude on my work and life.

I would like to thank my collaborators, Prof. Eyuphan Bulut, Prof. Chris Gaiteri and Prof. Jianxi Gao for their insightful guidance. Prof. Gao introduced me to the computational physics methods for complex network analysis. Prof. Bulut introduced me to work on scale-free P2P network construction. Prof. Chris Gaiteri worked with me on the collaboration recommendation system for biomedical researchers based on the co-authorship and molecule interaction networks. I appreciate the opportunity of working with Dr. Konstantin Kuzmin, Dr. Mingming Chen, and Robert Paluch during my doctoral study. I am grateful for the friendship and support of my fellow collaborators.

Besides my advisor, I would like to thank the rest of my thesis committee: Prof. Malik Magdon-Ismail, Prof. Gyorgy Korniss, and Prof. Jianxi Gao for kindly serving on my doctoral committee, and for their helpfulness and willingness to answer questions that arose during my time as a graduate student.

Finally, I would like to extend my love and gratitude to my family: Hui, Qiong and Xilei. I would not be where I am today without their love and support.
